Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-07-30 manuscripts/reply/reply-main.tex
More work on overhaul of intro to reply ms.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method was very sensitive to prior assumptions and often
incorrectly supported shared divergences across taxa, to which
\citet{Hickerson2013} responded.
Both papers agree on the fundamental methodological point about the model
selection performed in \msb:
% The papers of \citet{Oaks2012} and \citet{Hickerson2013} are in agreement
% on the fundamental methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       support for models with few divergence events shared across taxa. Thus,
       the primary inference enabled by the approach is very sensitive to the
       priors on divergence times.
\end{itemize}
However, the two papers suggest alternative causes for this behavior:
\begin{itemize}
    \item \citep{Hickerson2013} suggest broad uniform priors cause numerical
        problems in which the ABC rejection algorithm is unable to sufficiently
        sample the space of the models in reasonable computational time. The
        method will incorrectly support models with fewer divergence-time
        parameters because they are better sampled.
    \item \citet{Oaks2012} suggest the cause is more fundamental; the broad
        uniform priors simply lead to very small marginal likelihoods of models
        with many divergence-time parameters.
\end{itemize}
In the former case, the true posterior would support heterogeneity in 
divergences among taxa in the face of temporally random divergences; the
problem only arises from our inability to accurately estimate the posterior.
In the latter case, the true posterior would support simultaneous divergences
among taxa in the face of random divergences. I.e., under the prior
assumptions about divergence times, this is the correct answer from 
the perspective of Bayesian model choice.
Such posterior support for simultaneous divergence, even if correct, does not
provide the biogeographical insights that a researcher who employs \msb seeks
to gain.

The distinction between these potential causes is important for improving our
ability to estimate shared divergence histories.
If the cause is numerical, then the model is sound and we simply need to
improve our estimation procedures.
For example, \citet{Hickerson2013} proposed a model-averaging approach that
uses narrow, empirically informed uniform priors in the hope that with less
parameter space to sample, the rejection algorithm will produce better
estimates of the posterior.
If the cause is marginal likelihoods, we need to correct the model, because
improving the estimation machinery will only increase our ability to estimate
biogeographically misleading posteriors.
Accordingly, \citet{Oaks2014dpp} has introduced a method that uses more
flexible probability distributions to accommodate prior uncertainty in
divergence times without overly inhibiting the marginal likelihoods of models
with more divergence-time parameters, which increases the method's robustness
and power to detect temporal variation in divergences.

This is not surprising given the rich statistical literature showing that
marginal likelihoods are very sensitive to the priors used in Bayesian model
selection \citep[e.g.,][]{Jeffreys1939,Lindley1957}.
Thus, uniform priors on divergence times can impose a strong ``penalty'' for
divergence-time parameters, creating a strong preference for simple
models that are often of interest to biogeographers, even if these models do
not explain the data very well.

\change{
Here, we discuss theoretical considerations for using empirically informed
priors for Bayesian model choice.
}
We then evaluate the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} as a potential solution to the biases of \msb revealed by
\citet{Oaks2012}.
In their re-analysis of the dataset of \citet{Oaks2012}, \citet{Hickerson2013}
made an error by mixing different units of time, which makes the results
presented in their response difficult to interpret (see Supporting Information
for details).
We avoid this error, but still find their approach will often spuriously
support clustered divergence models.
Furthermore, the approach provides another means by which the method can
``escape'' models with large parameter space, which manifests in the preference
of models that exclude from consideration the true values of the models'
parameters.
\change{
Our results, show that it is difficult to choose a uniformly distributed prior
on divergence times that is broad enough to confidently contain the true
parameter values while being narrow enough to avoid spurious support for models
with less parameter space.
}
We also explore the predictions of the potential causes of the bias proposed by
\citet{Oaks2012} and \citet{Hickerson2013}.
We show the primary cause of the method's tendency to incorrectly infer shared
divergences is the uniform priors on divergence times leading to small marginal
likelihoods of models with more divergence-time parameters.
We also show that sampling error is not a major contributer to the problem.
Fortunately, \citet{Oaks2014dpp} has introduced a method that uses more
flexible probability distributions to accommodate prior uncertainty in
divergence times without overly inhibiting the marginal likelihoods of models
with more divergence-time parameters, which increases the method's robustness
and power to detect temporal variation in divergences.

% \msb \citep{Huang2011} uses approximate Bayesian computation (ABC) to estimate
% the distribution of divergence times among co-distributed pairs of taxa.
% It approximates a posterior probability over models that range from a single
% divergence-time parameter (i.e., simultaneous divergence of all pairs of taxa)
% to the fully generalized model in which each pair of taxa diverged at a
% unique time.
% % A full description of the model is given in \citet{Oaks2012}.
% % Frequently, the exact values of the divergence times are hard 
% % to determine because researchers typically lack precise knowledge 
% % of the rate at which substitutions occur.
% % The exact divergence time is therefore often viewed as a nuisance parameter,
% % and researchers focus on the number of divergence events.
% The papers of \citet{Oaks2012} and \citet{Hickerson2013} are in agreement
% on the fundamental methodological point about the model selection performed in \msb:
% \begin{itemize}
%    \item The use of vague priors on the divergence-time parameters can lead to
%        support for models with few divergence events shared across taxa. Thus,
%        the primary inference enabled by the approach is very sensitive to the
%        priors on divergence times.
% \end{itemize}
% This is not surprising given the rich statistical literature that shows that
% marginal likelihoods are very sensitive to the priors used in Bayesian model
% selection
% \citep[e.g.,][]{Jeffreys1939,Lindley1957}.
% Accordingly, \citet{Oaks2012} suggest the primary cause of spurious support
% for models with few divergence parameters is the greater marginal likelihoods
% of these models under vague uniform priors, relative to more parameter-rich
% models.

% \change{
% Models with more divergence-time parameters integrate over \emph{much} greater
% parameter space with low probability of producing the data, yet relatively high
% prior density imposed by the uniformly distributed prior.
% % By using such priors on divergence times, we are essentially telling the
% % method to favor models with few divergence times even if they do not
% % explain the data very well.
% Thus, uniform priors on divergence times can impose a strong ``penalty'' for
% divergence-time parameters, creating a strong preference for simple
% models that are often of interest to biogeographers, even if these models do
% not explain the data very well.
% It is important to note that this is not due to estimation error
% (i.e., this phenomenon is extrinsic to the numerical approximation machinery of
% \msb).
% Rather, the \emph{true} posterior will tend to support models with few
% divergence-time parameters under broad uniform priors on divergence times.
% Such posterior support for simultaneous divergence, even if correct from the
% perspective of Bayesian model choice, does not provide the biogeographical
% insights that a researcher who employs \msb seeks to gain.
% % Ideally, the estimated posterior probability of a simultaneous-divergence model
% % can be interpreted as an estimate of the probability that this model is correct.
% % However, when using uniform priors on divergence times, it might be better
% % interpreted as an estimate of the probability that the simultaneous-divergence
% % model cannot be completely ruled out as a possible explanation of the data.
% % Note, this is a statistical issue, extrinsic to \msb, and does not question the
% % soundness of the numerical approximation machinery of \msb.
% }

% \citet{Hickerson2013} suggest the problem is the result of a numerical issue
% intrinsic to \msb, concluding the method's rejection algorithm is inefficient
% and will be biased toward models with fewer divergence-time parameters, because
% the parameter space of these models are more densely sampled relative to models
% with more divergence-time parameters.
% We agree that sampling error is present in all numerical Bayesian estimation
% methods, however, we show here that it is not a major contributer to the
% biases found by \citet{Oaks2012}.
% % We present results here of analyses that, when combined with the results
% % of \citet{Oaks2012}, strongly suggest that \msb prefers models with few
% % divergence-time parameters, because they have greater likelihoods when
% % averaged over broad uniform priors on divergences times.

% \citet{Oaks2012} suggest alternative prior probability distributions on
% divergence-times could increase the marginal likelihoods of models with more
% divergence-time parameters and thus reduce spurious support for models of
% temporally clustered divergences.
% Alternatively, \citet{Hickerson2013} present an approach that uses empirically
% informed uniform priors and Bayesian model-averaging in an attempt to accommodate
% uncertainty in selecting priors.
% In general, we agree with the use of Bayesian model averaging to obtain a
% posterior estimate marginalized over uncertainty in prior choice.
% However, we question whether adding an additional dimension of model choice
% that samples over empirically informed uniform priors is a fruitful solution
% for a model-choice method that is highly sensitive to priors.

% \change{
% In this paper, we discuss theoretical and practical considerations of using
% empirically informed priors for Bayesian model choice.
% }
% Furthermore, we evaluate the empirical Bayesian model-averaging approach of
% \citet{Hickerson2013} as a potential solution to the biases of \msb revealed by
% \citet{Oaks2012}.
% In their re-analysis of the dataset of \citet{Oaks2012}, \citet{Hickerson2013}
% made an error by mixing different units of time, which makes the results
% presented in their response difficult to interpret (see Supporting Information
% for details).
% Here, we avoid this error, but still find their approach to be biased toward
% clustered divergence models.
% Furthermore, the approach provides another means by which the method can
% ``escape'' models with large parameter space, which manifests in the preference
% of models that exclude from consideration the true values of the model's
% parameters.
% \change{
% Our results, show that it is difficult to choose a uniformly distributed prior
% on divergence times that is broad enough to confidently contain the true values
% of parameters while being narrow enough to avoid spurious support of models
% with less parameter space.
% }
% However, recent work shows more flexible probability distributions without a
% hard upper bound (e.g., gamma) can accommodate prior uncertainty in divergence
% times without inhibiting the marginal likelihoods of models with more
% divergence-time parameters, and as a result, increase the method's power to
% detect temporal variation in divergences
% \citep{Oaks2014dpp}.

\section{The potential implications of empirical Bayesian model choice}
\change{
% The results of \citet{Oaks2012} and \citet{Hickerson2013} show the method is
% highly sensitive to the prior on divergence times.
\citet{Oaks2012} found \msb incorrectly supports models of temporally
clustered divergences even when using a prior on divergence times that was
informed by the data being analyzed.
The main argument of \citet{Hickerson2013} is that the priors used by
\citet{Oaks2012} were not informative enough.
}
%
% \citet{Hickerson2013} repeatedly refer to the prior distributions used by
% \citet{Oaks2012} for divergence-time parameters as ``poorly selected.''
% However, this is misleading, as \citet{Oaks2012} discuss in detail how they
% selected their prior to reflect the large amount of uncertainty about the
% timing of divergences across all 22 of the taxa in their study (see section
% ``Specifying and simulating the joint prior'' of \citet{Oaks2012}).
% \citet{Oaks2012} further discuss how the use of uniform prior distributions in
% \msb for many of the model's parameters requires investigators to select broad
% priors to avoid excluding the truth \textit{a priori}, resulting in too much
% density in improbable regions of parameter space.
%
% \citet{Hickerson2013} suggest that \citet{Oaks2012} should have used their
% data to inform their prior on divergence times (i.e., an empirical Bayesian
% approach).
% However, \citet{Oaks2012} did use empirically informed priors to mimic
% empirical situations in which a large amount of prior information is available;
% they did so to assess the prior sensitivity of the method and to determine if
% uniform priors in such situations are narrow enough to avoid spurious support
% of models with few divergence-time parameters.
% The results of \citet{Oaks2012} and \citet{Hickerson2013} show the method is
% highly sensitive to the prior on divergence times.
% Furthermore, \citet{Oaks2012} found the method remained biased toward clustered
% divergences under the empirically informed priors.
%
% The main argument of \citet{Hickerson2013} is that the priors used by
% \citet{Oaks2012} were not informative enough.
They suggest a very narrow, highly informed uniform prior on divergence times
is necessary to avoid the method's preference for models with few
divergence-time parameters.
Such an empirical Bayesian approach to model selection raises some theoretical
and practical concerns, some of which were discussed by \citet{Oaks2012} (see
the last paragraph of ``Assessing prior sensitivity of \msb'' in
\citet{Oaks2012}); we expand on this here.

>>>>

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method was very sensitive to prior assumptions and often
incorrectly supported shared divergences across taxa that diverged randomly
over broad time periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
% The papers of \citet{Oaks2012} and \citet{Hickerson2013} are in agreement
% on the fundamental methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       spurious support for models with few divergence events shared across
       taxa. Thus, the primary inference enabled by the approach is very
       sensitive to the priors on divergence times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{itemize}
    \item \citet{Hickerson2013} suggest that under broad uniform priors, the
        rejection algorithm implemented in \msb is unable to adequately sample
        the space of the models within reasonable computational time.
        They propose the method incorrectly supports models with fewer
        divergence-time parameters because they are better sampled.
    \item \citet{Oaks2012} suggest the cause is more fundamental; the broad
        uniform priors simply lead to very small marginal likelihoods of models
        with many divergence-time parameters.
        Models with more divergence-time parameters integrate over \emph{much}
        greater parameter space with low probability of producing the data, yet
        relatively high prior density, resulting in small marginal likelihoods.
\end{itemize}
In the former case, the true (correct) posterior supports models with many
divergence-time parameters given data generated under temporally random
divergences, but the problem arises from our inability to accurately estimate
the posterior.
In the latter case, estimation error is not the issue, because the true
posterior supports simultaneous divergences given data from randomly diverged
taxa; this is the correct (albeit biogeographically misleading) answer given
the prior assumptions about divergence times imposed by broad uniform priors.
Such posterior support for simultaneous divergence, even if correct from the
perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

While these phenomena are not mutually exclusive, it is important to
distinguish between them in order to determine how best to improve our ability
to estimate shared divergence histories.
If the cause is estimation error, then the model is sound and we need to
improve our numerical estimation procedures.
For example, \citet{Hickerson2013} proposed a model-averaging approach that
uses narrow, empirically informed uniform priors in the hope that with less
parameter space to sample, the rejection algorithm will produce better
estimates of the posterior.
If the cause is marginal likelihoods, we need to correct the model; improving
the estimation machinery will only increase our ability to estimate
biogeographically misleading posteriors.
Accordingly, \citet{Oaks2014dpp} has introduced a method that uses more
flexible probability distributions to accommodate prior uncertainty in
divergence times without overly inhibiting the marginal likelihoods of models
with more divergence-time parameters, which increases the method's robustness
and power to detect temporal variation in divergences.

Here, we discuss theoretical considerations for using empirically informed
priors for Bayesian model choice and evaluate the empirical Bayesian
model-averaging approach of \citet{Hickerson2013} as a potential solution to
the biases of \msb revealed by \citet{Oaks2012}.
In their re-analysis of the dataset of \citet{Oaks2012}, \citet{Hickerson2013}
made an error by mixing different units of time, which invalidates the results
presented in their response (see Supporting Information for details).
We avoid this error, but find their approach will often incorrectly support
(1) clustered divergence models and
(2) models that exclude from consideration the true values of the models'
parameters.
% Our results show that it is difficult to choose a uniformly distributed prior
% on divergence times that is broad enough to confidently contain the true
% parameter values while being narrow enough to avoid spurious support for models
% with less parameter space.
We also explore the distinct predictions made by the alternative causes of the
bias proposed by \citet{Oaks2012} and \citet{Hickerson2013}.
We show the primary cause of the method's tendency to incorrectly infer shared
divergences is the uniform priors on divergence times leading to small marginal
likelihoods of models with more divergence-time parameters.
We also show that sampling error is not a major contributer to the problem.
This is not surprising given the rich statistical literature showing that
marginal likelihoods are very sensitive to the priors used in Bayesian model
selection
\citep[e.g.,][]{Jeffreys1939,Lindley1957}.
As predicted by this fundamental principle of Bayesian model selection,
alternative prior assumptions can improve robustness of the method and greatly
mitigate misleading support for models of shared divergences
\citet{Oaks2014dpp}.

\section{The potential implications of empirical Bayesian model choice}
\change{
\citet{Oaks2012} found \msb incorrectly supports models of temporally
clustered divergences even when using a prior on divergence times that was
informed by the data being analyzed.
The main argument of \citet{Hickerson2013} is that the priors used by
\citet{Oaks2012} were not informative enough.
}
%
They suggest a very narrow, highly informed uniform prior on divergence times
is necessary to avoid the method's preference for models with few
divergence-time parameters.
Such an empirical Bayesian approach to model selection raises some theoretical
and practical concerns, some of which were discussed by \citet{Oaks2012} (see
the last paragraph of ``Assessing prior sensitivity of \msb'' in
\citet{Oaks2012}); we expand on this here.
